---
title: "LDA implementations in R and Python"
subtitle: "Where tidylda sits in the ecosystem"
output:
  html_document:
    theme: null
    css: null
    toc: true
    toc_float: true
    df_print: kable
    fig_width: 10
    fig_height: 6
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dpi = 130)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

root <- Sys.getenv("LDA_BENCH_ROOT", unset = path.expand("~/lda-benchmarks"))
source(file.path(root, "R", "grid.R"))
source(file.path(root, "R", "theme.R"))

runs <- read.csv(file.path(root, "results", "runs.csv"), stringsAsFactors = FALSE)

families <- vapply(names(ENGINES), function(e) ENGINES[[e]]$family, character(1))
runs$family <- unname(families[runs$engine])
runs$family <- factor(runs$family, levels = names(FAMILY_COLORS))
runs$corpus <- factor(runs$corpus, levels = c("ap", "20ng"),
                      labels = c("AP (2.2k docs)", "20 Newsgroups (17.7k docs)"))

# Reps are collapsed by median: with three runs the median is the robust choice
# and it is what the timing numbers in the text refer to.
agg <- runs %>%
  group_by(block, engine, family, corpus, k, iters, threads) %>%
  summarise(
    n_reps    = dplyr::n(),
    fit_sec   = median(fit_sec),
    wall_sec  = median(wall_sec),
    max_rss_mb = median(max_rss_kb) / 1024,
    r2        = median(r2),
    coherence = median(coherence_mean),
    .groups = "drop"
  )

# Label only the end of each curve, never every point.
end_labels <- function(d, xvar, yvar) {
  d %>% group_by(engine, corpus) %>%
    slice_max(order_by = .data[[xvar]], n = 1, with_ties = FALSE) %>%
    ungroup()
}
```

## What this measures

Ten LDA implementations, fit on identical document-term matrices with identical
hyperparameters (`alpha = 0.1`, `eta = 0.05`, no hyperparameter optimization anywhere), scored
by one common code path. Each run is its own OS process under `/usr/bin/time -v`.

Color encodes the algorithm family. Engine names are labeled directly on the curves.

```{r corpora}
manifests <- lapply(c("ap", "20ng"), function(cp) {
  m <- jsonlite::fromJSON(file.path(root, "data", cp, "manifest.json"))
  data.frame(Corpus = cp, Documents = m$n_docs, Terms = m$n_terms,
             Tokens = m$n_tokens, `Mean doc length` = round(m$mean_doc_len, 1),
             check.names = FALSE)
})
knitr::kable(do.call(rbind, manifests), format.args = list(big.mark = ","),
             caption = "Corpora")
```

## 1. Quality against wall-clock time

The headline result. A Gibbs sweep and a variational pass are different units of work, so
comparing "200 iterations each" would compare two arbitrary numbers. Instead each engine is run
at a ladder of iteration counts, single threaded, and we plot the quality it reached against how
long it took. **Down and to the right is worse; up and to the left is better.** A curve that sits
above and to the left of another dominates it.

```{r frontier-coherence, fig.height = 5.5}
d <- agg %>% filter(block == "frontier", threads == 1)

ggplot(d, aes(fit_sec, coherence, color = family, group = engine)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  geom_point(size = 2.2) +
  ggrepel::geom_text_repel(
    data = end_labels(d, "fit_sec", "coherence"),
    aes(label = engine), size = 3.1, hjust = 0, direction = "y",
    nudge_x = 0.08, segment.color = INK_MUTED, segment.size = 0.25,
    min.segment.length = 0, show.legend = FALSE
  ) +
  scale_x_log10() +
  scale_color_manual(values = FAMILY_COLORS, name = NULL) +
  facet_wrap(~corpus, scales = "free") +
  labs(
    title = "Mean probabilistic coherence vs. fit time",
    subtitle = "k = 100, single threaded. Each point is one iteration-ladder setting.",
    x = "Fit time (seconds, log scale)", y = "Mean probabilistic coherence",
    caption = "Median of replicate runs. Coherence computed by tidylda::calc_prob_coherence(m = 5) on the training DTM."
  ) +
  theme_bench() +
  expand_limits(x = max(d$fit_sec) * 1.6)
```

```{r frontier-r2, fig.height = 5.5}
ggplot(d, aes(fit_sec, r2, color = family, group = engine)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  geom_point(size = 2.2) +
  ggrepel::geom_text_repel(
    data = end_labels(d, "fit_sec", "r2"),
    aes(label = engine), size = 3.1, hjust = 0, direction = "y",
    nudge_x = 0.08, segment.color = INK_MUTED, segment.size = 0.25,
    min.segment.length = 0, show.legend = FALSE
  ) +
  scale_x_log10() +
  scale_color_manual(values = FAMILY_COLORS, name = NULL) +
  facet_wrap(~corpus, scales = "free") +
  labs(
    title = "R-squared vs. fit time",
    subtitle = "k = 100, single threaded.",
    x = "Fit time (seconds, log scale)", y = expression(R^2),
    caption = "In-sample, via mvrsquared::calc_rsquared on the training DTM."
  ) +
  theme_bench() +
  expand_limits(x = max(d$fit_sec) * 1.6)
```

## 2. Thread scaling

Only some engines thread at all. For those that do, this is speedup relative to their own
single-threaded time — so it answers "does adding cores help *this* package", not "which package
is fastest" (that is the chart above).

```{r thread-speedup, fig.height = 6}
base <- agg %>%
  filter(block == "frontier", threads == 1) %>%
  select(engine, corpus, k, iters, base_sec = fit_sec)

sc <- agg %>%
  filter(block %in% c("threads", "frontier")) %>%
  inner_join(base, by = c("engine", "corpus", "k", "iters")) %>%
  mutate(speedup = base_sec / fit_sec)

if (nrow(sc)) {
  ggplot(sc, aes(threads, speedup, color = family)) +
    geom_abline(slope = 1, intercept = 0, linetype = "22",
                color = INK_MUTED, linewidth = 0.4) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2) +
    facet_grid(corpus ~ engine) +
    scale_color_manual(values = FAMILY_COLORS, name = NULL) +
    scale_x_continuous(breaks = THREAD_LEVELS) +
    labs(
      title = "Speedup from additional cores",
      subtitle = "Relative to each engine's own single-threaded time. Dashed line is linear scaling.",
      x = "Threads", y = "Speedup",
      caption = "Engines absent from this panel are single threaded by construction."
    ) +
    theme_bench()
}
```

Threading is not free of consequences for every engine. MALLET and tomotopy use parallel samplers
that do not preserve the exact collapsed Gibbs chain; tidylda's is documented to be invariant to
thread count, and `scripts/98-verify.R` confirms its parameters are bit-identical at 1 and 20
threads. So it is worth checking whether quality moves as cores are added:

```{r thread-quality, fig.height = 5}
if (nrow(sc)) {
  ggplot(sc, aes(threads, coherence, color = family, group = engine)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2) +
    facet_grid(corpus ~ engine, scales = "free_y") +
    scale_color_manual(values = FAMILY_COLORS, name = NULL) +
    scale_x_continuous(breaks = THREAD_LEVELS) +
    labs(
      title = "Does quality change with thread count?",
      subtitle = "Flat is what an exact sampler looks like.",
      x = "Threads", y = "Mean coherence"
    ) +
    theme_bench()
}
```

## 3. Memory

Peak resident set size of the whole process, so it includes the interpreter or JVM baseline.
The baselines are shown rather than subtracted.

```{r memory, fig.height = 5}
mem <- agg %>%
  filter(block == "frontier", threads == 1, k == DEFAULT_K) %>%
  group_by(engine, family, corpus) %>%
  slice_max(order_by = iters, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot(mem, aes(reorder(engine, max_rss_mb), max_rss_mb, fill = family)) +
  geom_col(width = 0.7) +
  coord_flip() +
  facet_wrap(~corpus, scales = "free_x") +
  scale_fill_manual(values = FAMILY_COLORS, name = NULL) +
  labs(
    title = "Peak memory",
    subtitle = "Whole-process peak RSS at the longest iteration setting, k = 100, single threaded.",
    x = NULL, y = "Peak RSS (MB)"
  ) +
  theme_bench()
```

## 4. Quality across topic counts

```{r ksweep, fig.height = 5}
ks <- agg %>%
  filter(corpus == levels(agg$corpus)[1], threads == 1,
         block %in% c("k", "frontier")) %>%
  filter(iters == vapply(engine, engine_fixed_iters, numeric(1)))

if (nrow(ks)) {
  ks %>%
    pivot_longer(c(coherence, r2), names_to = "metric", values_to = "value") %>%
    mutate(metric = recode(metric, coherence = "Mean coherence", r2 = "R-squared")) %>%
    ggplot(aes(k, value, color = family, group = engine)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2) +
    ggrepel::geom_text_repel(
      data = function(d) d %>% group_by(engine, metric) %>%
        slice_max(k, n = 1, with_ties = FALSE) %>% ungroup(),
      aes(label = engine), size = 3, hjust = 0, direction = "y",
      nudge_x = 8, segment.color = INK_MUTED, segment.size = 0.25,
      min.segment.length = 0, show.legend = FALSE
    ) +
    facet_wrap(~metric, scales = "free_y") +
    scale_color_manual(values = FAMILY_COLORS, name = NULL) +
    scale_x_continuous(breaks = K_LEVELS) +
    labs(title = "Quality across topic counts", subtitle = "AP, single threaded.",
         x = "Number of topics (k)", y = NULL) +
    theme_bench() +
    expand_limits(x = max(K_LEVELS) * 1.35)
}
```

## 5. The numbers

Identity is never carried by color alone: here is every aggregated run.

```{r table}
agg %>%
  filter(block == "frontier", threads == 1) %>%
  transmute(Corpus = corpus, Engine = engine, Family = family, k, Iterations = iters,
            `Fit (s)` = round(fit_sec, 2),
            `Peak RSS (MB)` = round(max_rss_mb),
            `R^2` = round(r2, 4),
            Coherence = round(coherence, 4),
            Reps = n_reps) %>%
  arrange(Corpus, Engine, Iterations) %>%
  knitr::kable()
```

## Environment

```{r env, comment = ""}
cat(readLines(file.path(root, "results", "env", "hardware.txt")), sep = "\n")
cat("\n")
cat(readLines(file.path(root, "results", "env", "tidylda-provenance.txt")), sep = "\n")
```

```{r versions}
runs %>%
  distinct(engine, language, version) %>%
  arrange(language, engine) %>%
  rename(Engine = engine, Language = language, Version = version) %>%
  knitr::kable(caption = "Implementation versions")
```
